---
title: "earthquakes"
author: "Sarah"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(dplyr)
library(janitor)
library(corrplot)
library(GGally)
library(lubridate)
library(skimr)
library(geosphere)

```


#Merging Data Sets
added 1 column to the static readers of "user_Id" so I could rbind data sets
```{r}
mobile_read <- read_csv("MobileSensorReadings.csv") %>% 
  clean_names() 

# %>%  mutate(u_id = paste(sensor_id, timestamp, sep = "_"))

names(mobile_read)

static_read <- read_csv("StaticSensorReadings.csv") %>% 
  clean_names()

static_loc <- read_csv("StaticSensorLocations.csv") %>% 
  clean_names()

#joing static longitude and latitude
stat_read_loc <- left_join(static_read,static_loc,by="sensor_id")

# reorder static data and add user_id column
stat_read_loc <- stat_read_loc %>% 
  select(timestamp,sensor_id,long,lat,value,units) %>% 
  mutate(user_id = "static_sensor")

```
## long data
```{r}
# combine data sets
all_data <- rbind(mobile_read,stat_read_loc)
```


### updating for usefull date features
```{r}
all_data <- all_data %>%
  mutate(
    hour = hour(timestamp),
    date = as.Date(timestamp),
    weekday = wday(timestamp, label = TRUE),
    sensor_type = ifelse(user_id =="static_sensor","static","mobile")
  )

all_data <- all_data %>%
  mutate(
    sensor_id = as.factor(sensor_id),
    user_id = as.factor(user_id),
    units = as.factor(units),
    weekday = as.factor(weekday),
    sensor_type = as.factor(sensor_type)
  ) 


```

## omit 1 large value outlire
```{r}
no_outlier <- all_data %>% 
  filter(value != max(value, na.rm = TRUE))
```
# Data exploration
## Data Set Overview
```{r}
all_data %>%
  group_by(sensor_type) %>%
  skim()
```
```{r}
ggplot(no_outlier, aes(x = value, fill = sensor_type)) + 
  geom_histogram(position = "dodge", bins = 30) + 
  facet_wrap(~ sensor_type, scales = "free") +
  theme_minimal() +
  labs(title = "Histogram of Value Stratified by Sensor Type", 
       x = "Value", y = "Count") 
```

```{r}
boxplot(all_data$value, main = "Boxplot of Values", col = "lightblue", horizontal = TRUE)
```
## user counts
```{r}
user_counts <- all_data %>%
  group_by(user_id) %>%
  summarise(count = n())

ggplot(user_counts, aes(x = user_id, y = count)) +
  geom_bar(stat = "identity", fill = "lightblue") +
  labs(
    title = "Submission Count by User",
    x = "",
    y = "Number of Records"
  ) +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 1))
```


##Data Review with outlier values
```{r}
skim(all_data)
```

```{r}
summary(all_data)
str(all_data)
dim(all_data)
```

## Data review without outlier value
```{r}
skim(no_outlier)
```
```{r}
summary(no_outlier)
str(no_outlier)
dim(no_outlier)
```
## Data exploration without outlier

### Distribution
```{r}
# Distribution of `value`
hist(no_outlier$value, main="Distribution of Value", xlab="Value", col="skyblue")

# Distribution of `hour`
hist(no_outlier$hour, main="Distribution of Hour", xlab="Hour of Day", col="lightgreen")
```


### Geospatial scatter plot
```{r}
# If `long` and `lat` are geospatial, use a scatter plot
plot(no_outlier$long, no_outlier$lat, main="Geospatial Distribution", xlab="Longitude", ylab="Latitude", col="darkred", pch=20)
```

## Group-by for sensors
### Date Group-by
```{r}
# Aggregate by date
daily_data <- no_outlier %>%
  group_by(date) %>%
  summarise(
    avg_value = mean(value, na.rm = TRUE),
    min_value = min(value, na.rm = TRUE),
    max_value = max(value, na.rm = TRUE),
    count = n()
  )

# Plot the daily average values
plot(daily_data$date, daily_data$avg_value, type="l", col="blue", main="Daily Average Value", xlab="Date", ylab="Average Value")

```
### Weekday Group-by
```{r}
# Aggregate by weekday
weekday_data <- no_outlier %>%
  filter(sensor_type != "mobile") %>% 
  group_by(weekday) %>%
  summarise(
    avg_value = mean(value, na.rm = TRUE),
    min_value = min(value, na.rm = TRUE),
    max_value = max(value, na.rm = TRUE)
  )

# Plot weekday distribution of values
barplot(weekday_data$avg_value, names.arg = weekday_data$weekday, main = "Average Value by Weekday [static]", col = "pink", xlab = "Weekday", ylab = "Average Value")

```

### Group-by Sensor ID
```{r}
# Aggregate by sensor_id
sensor_data <- no_outlier %>%
  group_by(sensor_type, sensor_id) %>%
  summarise(
    avg_value = mean(value, na.rm = TRUE),
    min_value = min(value, na.rm = TRUE),
    max_value = max(value, na.rm = TRUE),
    count = n()
  )


# Plot with ggplot2
ggplot(sensor_data, aes(x = factor(sensor_id), y = avg_value, fill = sensor_type)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Average Value by Sensor and Type",
    x = "Sensor ID",
    y = "Average Value",
    fill = "Sensor Type"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5))
```
### Group-by User ID
```{r}
no_outlier %>%
  group_by(user_id) %>%
  summarise(
    avg_value = mean(value, na.rm = TRUE),
    min_value = min(value, na.rm = TRUE),
    max_value = max(value, na.rm = TRUE),
    count = n()
  ) %>% 


# Plot sensor ID-based value distribution
ggplot(aes(x = as.factor(user_id), y = avg_value)) +
  geom_bar(stat = "identity", fill = "lightgreen") +
  labs(
    title = "Average Value by User",
    x = "",  # Removes "User ID" label
    y = "Average Value"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```


## Correlation between numeric Variables
```{r}
# Correlation matrix for numeric columns (value, long, lat, hour)
cor_data <- no_outlier %>% select(value, long, lat, hour)
cor_matrix <- cor(cor_data, use = "complete.obs")
print(cor_matrix)

# Plot correlation heatmap
library(ggplot2)
library(reshape2)

cor_melted <- melt(cor_matrix)
ggplot(cor_melted, aes(Var1, Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0) +
  theme_minimal() +
  labs(title = "Correlation Heatmap", x = "", y = "") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```

### Geospatial
```{r}
# Install and load necessary packages
library(ggplot2)
library(ggmap)

# Plot the geospatial data
ggplot(no_outlier, aes(x = long, y = lat)) +
  geom_point(aes(color = value), size = 0.5, alpha = 0.7) +
  labs(title = "Geospatial Distribution of Values", x = "Longitude", y = "Latitude") +
  scale_color_viridis_c()
 
```

## Timeseries
### Aggregated at each Timestamp
Aggregated the data by timestamp and take the average value at each time point

```{r}
# Aggregate the data by timestamp (calculate the average value for each timestamp)
aggregated_data <- no_outlier %>%
  group_by(sensor_type,timestamp) %>%
  summarise(avg_value = mean(value, na.rm = TRUE), .groups = "drop")

# Plot the time series with sensor_type distinction
ggplot(aggregated_data, aes(x = timestamp, y = avg_value, color = sensor_type)) +
  geom_line() +
  labs(
    title = "Aggregated Time Series of Values by Sensor Type",
    x = "Time",
    y = "Average Value",
    color = "Sensor Type"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```
### Median Time Series: 
preserve the distribution of values calculated the median each timestamp
```{r}
# Aggregate using median instead of mean
aggregated_data_median <- no_outlier %>%
  group_by(sensor_type,timestamp) %>%
  summarise(median_value = median(value, na.rm = TRUE), .groups = "drop")





# Plot the time series with sensor_type distinction
ggplot(aggregated_data_median, aes(x = timestamp, y = median_value, color = sensor_type)) +
  geom_line() +
  labs(
    title = "Aggregated Time Series of Values by Sensor Type",
    x = "Time",
    y = "Median Value",
    color = "Sensor Type"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```



```{r}
library(zoo)

# Compute rolling average (moving window of 24 hours)
no_outlier <- no_outlier %>%
  arrange(timestamp) %>%
  mutate(rolling_avg = rollmean(value, 24, fill = NA, align = "right"))

# Plot rolling average over time
ggplot(no_outlier, aes(x = timestamp, y = rolling_avg)) +
  geom_line(color = "blue") +
  labs(title = "Rolling Average of Value (24 Hours)", x = "Time", y = "Rolling Average Value")
  
```

# Project Task Analysis
## Examine Spread of Contamination over Time

```{r}
model_weekday <- lmer(value ~ weekday + sensor_type + (1 | sensor_type), data = no_outlier)
summary(model_weekday)

```


### Static versus Mobile
```{r}
# Aggregate contamination levels over time by sensor type, sensor_id, and region
radiation_over_time <- no_outlier %>%
  group_by(timestamp, lat, long, sensor_id, sensor_type) %>%  # Group by sensor_type
  summarise(avg_value = mean(value, na.rm = TRUE), .groups = "drop")

# Plotting radiation levels over time, broken down by sensor type
ggplot(radiation_over_time, aes(x = timestamp, y = avg_value, color = sensor_type)) +
  geom_line() +
  facet_wrap(~sensor_type, scales = "free_y") +  # Separate the plot by sensor_type
  labs(
    title = "Radiation Levels Over Time by Sensor Type",
    x = "Time",
    y = "Average Radiation Value"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotat
```


```{r}
ggplot(no_outlier, aes(x = long, y = lat, color = value, alpha = value)) +
  geom_point(size = 2) +
  facet_wrap(~date) +
  scale_alpha_continuous(range = c(0, 1), trans = "identity") +
  scale_color_gradient(
    low = "lightyellow",  # low value color
    high = "darkred"      # high value color (dark!)
  ) +
  labs(
    title = "CPM Value for Mobile Sensors by Region",
    x = "Longitude",
    y = "Latitude",
    color = "Value"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


```{r}

# 1. Create bins for location
no_outlier <- no_outlier %>%
  mutate(
    lon_bin = round(long, 2),  # Adjust 2 for finer or coarser grid
    lat_bin = round(lat, 2)
  )

# 2. Aggregate by location bin + time
spatial_temporal <- no_outlier %>%
  filter(sensor_type == "mobile") %>% 
  group_by(lon_bin, lat_bin, timestamp) %>%
  summarise(avg_value = mean(value, na.rm = TRUE), .groups = 'drop')

# 3. Plot heatmap for a specific time (e.g., pick one timestamp)
selected_time <- as.POSIXct("2020-04-06 00:00:05")  # example time

heatmap_data <- spatial_temporal %>%
  filter(timestamp == selected_time)

ggplot(heatmap_data, aes(x = lon_bin, y = lat_bin, fill = avg_value)) +
  geom_tile() +
  scale_fill_viridis_c() +
  labs(title = paste("Radiation Map at", selected_time),
       x = "Longitude", y = "Latitude", fill = "Radiation Level") +
  coord_fixed() +
  theme_minimal()
```

```{r}
# 2. Aggregate by location bin + time
spatial_temporal2 <- no_outlier %>%
  filter(sensor_type != "mobile") %>% 
  group_by(lon_bin, lat_bin, timestamp) %>%
  summarise(avg_value = mean(value, na.rm = TRUE), .groups = 'drop')

# 3. Plot heatmap for a specific time (e.g., pick one timestamp)
selected_time <- as.POSIXct("2020-04-06 00:00:05")  # example time

heatmap_data2 <- spatial_temporal2 %>%
  filter(timestamp == selected_time)

ggplot(heatmap_data2, aes(x = lon_bin, y = lat_bin, fill = avg_value)) +
  geom_tile() +
  scale_fill_viridis_c() +
  labs(title = paste("Radiation Map at", selected_time),
       x = "Longitude", y = "Latitude", fill = "Radiation Level") +
  coord_fixed() +
  theme_minimal()
```


```{r}
# Install required packages if you haven't
# install.packages("gganimate")
# install.packages("viridis")

library(ggplot2)
library(gganimate)
library(viridis)

# 2. Aggregate by location bin + time + sensor_type
spatial_temporal <- no_outlier %>%
  group_by(sensor_type, lon_bin, lat_bin, timestamp) %>%
  summarise(avg_value = mean(value, na.rm = TRUE), .groups = 'drop')

# 3. Create the animated plot
animated_plot <- ggplot(spatial_temporal, aes(x = lon_bin, y = lat_bin, fill = avg_value)) +
  geom_tile() +
  scale_fill_viridis_c(direction = -1) +  # Darker colors = higher value
  labs(
    title = "Radiation Map",
    x = "Longitude",
    y = "Latitude",
    fill = "Radiation Level"
  ) +
  facet_wrap(~sensor_type) +  # Facet by sensor_type
  coord_fixed() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  transition_time(as.numeric(timestamp)) 

# Render the animation
animate(animated_plot, nframes = 100, fps = 10, width = 800, height = 600)

anim_save("radiation_animation.gif", animated_plot)
```
```{r}
library(Matrix)
library(lme4)
model_weekday <- lmer(value ~ weekday + sensor_type + (1 | sensor_type), data = no_outlier)
summary(model_weekday)
```



## Uncertainty Analysis
### # Step 1: Calculate standard deviation (SD) as uncertainty for the entire dataset 
(forcing 0's where there is only 1 observation)
```{r}

uncertainty_analysis <- no_outlier %>%
  group_by(sensor_type, date, lat, long) %>%
  summarise(
    avg_value = mean(value, na.rm = TRUE),
    sd_value = ifelse(n() > 1, sd(value, na.rm = TRUE), 0),  # Assign 0 if only one observation
    .groups = "drop"
  )

daily_data <- no_outlier %>%
  group_by(date, sensor_type) %>%
  summarise(daily_value = mean(value, na.rm = TRUE),
            daily_sd = sd(value,na.rm = TRUE))  # Or use sum() if more appropriate
```

### Uncertainty by day heatmap
```{r}
# Compute the uncertainty (SD) by date and weekday
uncertainty_heatmap <- no_outlier %>%
  group_by(date, weekday) %>%
  summarise(
    avg_sd_value = sd(value, na.rm = TRUE),
    .groups = "drop"
  )

# Plot the uncertainty heatmap
ggplot(uncertainty_heatmap, aes(x = date, y = weekday, fill = avg_sd_value)) +
  geom_tile(color = "white") +  # Color the tiles based on the SD value
  scale_fill_gradient(low = "white", high = "red") +  # Define color gradient for uncertainty
  labs(
    title = "Uncertainty (SD) Heatmap by Date and Weekday",
    x = "Date",
    y = "Day of the Week",
    fill = "Average SD (Uncertainty)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
```


### Step 2: Separate mobile and static sensor data
```{r}
mobile_data <- uncertainty_analysis %>% filter(sensor_type == "mobile")
static_data <- uncertainty_analysis %>% filter(sensor_type == "static")
```




###Plt SD (for mobile)
```{r}
ggplot(mobile_data, aes(x = long, y = lat, color = sd_value, alpha = sd_value)) +
  geom_point(size = 2) +  # Adjust size for visibility
  facet_wrap(~date) +
  scale_alpha_continuous(range = c(0, 1), trans = "identity") +  # Make 0 fully transparent
  labs(
    title = "Uncertainty (SD) for Mobile Sensors by Region",
    x = "Longitude",
    y = "Latitude",
    color = "Standard Deviation (Uncertainty)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
```

### Plt sd (Static)
```{r}
ggplot(static_data, aes(x = long, y = lat, color = sd_value, alpha = sd_value)) +
  geom_point(size = 2) +  # Adjust size for visibility
  facet_wrap(~date) +
  scale_alpha_continuous(range = c(0, 1), trans = "identity") +  # Make 0 fully transparent
  labs(
    title = "Uncertainty (SD) for Static Sensors by Region",
    x = "Longitude",
    y = "Latitude",
    color = "Standard Deviation (Uncertainty)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
```
```{r}
ggplot(daily_data, aes(x = date, y = daily_sd, color = sensor_type)) +
  geom_line() +
  labs(
    title = "Daily Radiation SD by Sensor Type",
    x = "",
    y = "Average Value",
    color = "Sensor Type"
  ) +
  theme_minimal()
```
### Test analysis between static and mobile SD's 
```{r}
t.test(daily_data$daily_sd~daily_data$sensor_type)
wilcox.test(daily_sd ~ sensor_type, data = daily_data)
```
```{r}
ggplot(daily_data, aes(x = sensor_type, y = daily_sd, fill = sensor_type)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.5, color = "black", size = 1.5) +
  stat_summary(fun = mean, geom = "point", shape = 18, size = 4, color = "red") +
  labs(title = "Daily Standard Deviation by Sensor Type",
       y = "Daily Standard Deviation", x = "Sensor Type") +
  theme_minimal() +
  theme(legend.position = "none")
```


### Test analysis between static and mobile SD's 
```{r}
t.test(daily_data$daily_value~daily_data$sensor_type)
wilcox.test(daily_value ~ sensor_type, data = daily_data)
```

Mean analysis
```{r}
ggplot(daily_data, aes(x = date, y = daily_value, color = sensor_type)) +
  geom_line() +
  labs(
    title = "Daily Radiation Values by Sensor Type",
    x = "",
    y = "Average Value",
    color = "Sensor Type"
  ) +
  theme_minimal()
```



```{r}
shapiro.test(daily_data$daily_value[daily_data$sensor_type == "mobile"])
shapiro.test(daily_data$daily_value[daily_data$sensor_type == "static"])
```
Mobile sensors definitely have higher daily radiation levels than static (strong evidence).
 Mobile sensors probably have higher variability than static, but you don't have strong enough evidence yet. (larger sample or different analysis might help.)
 We find strong evidence that mobile sensors record higher daily radiation levels than static sensors (p = 0.008), and some weak evidence that mobile sensors also have greater day-to-day variability (p = 0.095).
 
 

```{r}
ggplot(daily_data, aes(x = sensor_type, y = daily_value, fill = sensor_type)) +
  geom_boxplot() +
  labs(title = "Comparison of Daily Sensor Values: Mobile vs Static")
```
```{r}
ggplot(daily_data, aes(x = sensor_type, y = daily_sd, fill = sensor_type)) +
  geom_boxplot() +
  labs(title = "Comparison of Daily Sensor SD: Mobile vs Static")
```


###Plt SD (for mobile)
```{r}
ggplot(mobile_data, aes(x = long, y = lat, color = avg_value, alpha = avg_value)) +
  geom_point(size = 2) +
  facet_wrap(~date) +
  scale_alpha_continuous(range = c(0, 1), trans = "identity") +
  scale_color_gradient(
    low = "yellow",   # light color for small values
    high = "darkred"  # dark color for large values
  ) +
  labs(
    title = "Average Value for Mobile Sensors by Region",
    x = "Longitude",
    y = "Latitude",
    color = "Average Radiation Value"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

### Plt sd (Static)
```{r}
ggplot(static_data, aes(x = long, y = lat, color = avg_value, alpha = avg_value)) +
  geom_point(size = 2) +
  facet_wrap(~date) +
  scale_alpha_continuous(range = c(0, 1), trans = "identity") +
  scale_color_gradient(
    low = "yellow",   # light color for small values
    high = "darkred"  # dark color for large values
  ) +
  labs(
    title = "Average Value for Static Sensors by Region",
    x = "Longitude",
    y = "Latitude",
    color = "Average Radiation Value"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

```{r}
# Load libraries
library(sf) 
library(dplyr)
library(ggplot2)

# Assume your data is in a data frame called `df` with columns: longitude, latitude
# Convert to sf object
points_sf <- st_as_sf(no_outlier, coords = c("long", "lat"), crs = 4326)  # WGS 84

# Transform to a projected CRS for accurate distance-based binning (e.g., UTM or local)
points_proj <- st_transform(points_sf, crs = 3857)  # Web Mercator (good for small areas)

# Create a grid over the city area
grid <- st_make_grid(points_proj, cellsize = 500, square = TRUE)  # 500m x 500m grid
grid_sf <- st_sf(grid_id = 1:length(grid), geometry = grid)

# Spatial join: assign each point to a grid cell
points_with_grid <- st_join(points_proj, grid_sf)

# Count points per grid cell
grid_counts <- points_with_grid %>%
  st_drop_geometry() %>%
  group_by(grid_id) %>%
  summarise(count = n()) %>%
  left_join(grid_sf, by = "grid_id") %>%
  st_as_sf()

# Plot (optional)
ggplot(grid_counts) +
  geom_sf(aes(fill = count)) +
  scale_fill_viridis_c() +
  theme_minimal()
```

```{r}
# Drop geometry to simplify aggregation
agg_data <- points_with_grid %>%
  st_drop_geometry() %>%
  group_by(grid_id, sensor_type) %>%
  summarise(mean_value = mean(value, na.rm = TRUE),
            count = n(), .groups = "drop")

# Join back to grid geometry
agg_grid <- agg_data %>%
  left_join(grid_sf, by = "grid_id") %>%
  st_as_sf()
```

```{r}
# Calculate threshold (you can adjust this logic)
threshold <- quantile(agg_data$mean_value, 0.75, na.rm = TRUE)

# Flag trouble areas
agg_grid <- agg_grid %>%
  mutate(trouble_area = mean_value > threshold)
```

```{r}
ggplot(agg_grid) +
  geom_sf(aes(fill = trouble_area), color = NA) +
  facet_wrap(~ sensor_type) +
  scale_fill_manual(values = c("FALSE" = "lightgray", "TRUE" = "red")) +
  labs(
    title = "Identified Trouble Areas by Sensor Type",
    fill = "Trouble Area"
  ) +
  theme_minimal()

```

